---
title: "Linear Regression"
author: "Art Tay"
output: pdf_document
---

```{r setup, include=FALSE}
##Setup code
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# Libraries
library(tidyverse)
```

## Gradient Derivation
a)
\begin{align*}
\dfrac{\partial}{\partial \beta_0}
\left[
\frac{1}{2m} \sum_{i = 0}^{m}
(\beta_0 + \beta_1 x^{(i)} - y^{(i)})^2
\right]
&=
\dfrac{\partial}{\partial \beta_0}
\left[
\frac{1}{2m} \sum_{i = 0}^{m}
(\beta_0 + \beta_1 x^{(i)} - y^{(i)})^2
\right]
\\
&=
\frac{1}{2m} \sum_{i = 0}^{m}
\left[
2(\beta_0 + \beta_1 x^{(i)} - y^{(i)}) \cdot 1
\right]
\\
&=
\frac{1}{m} \sum_{i = 0}^{m} \beta_0 +
\beta_1 \frac{1}{m} \sum_{i = 0}^{m} x^{(i)} -
\frac{1}{m} \sum_{i = 0}^{m} y^{(i)}
\\
&=
\beta_0 + \beta_1 \bar x - \bar y
\end{align*}
b)
\begin{align*}
\dfrac{\partial}{\partial \beta_1}
\left[
\frac{1}{2m} \sum_{i = 0}^{m}
(\beta_0 + \beta_1 x^{(i)} - y^{(i)})^2
\right]
&=
\frac{1}{2m} \sum_{i = 0}^{m}
\dfrac{\partial}{\partial \beta_1}
\left[
(\beta_0 + \beta_1 x^{(i)} - y^{(i)})^2
\right]
\\
&=
\frac{1}{2m} \sum_{i = 0}^{m}
\left[
2 x^{(i)}(\beta_0 + \beta_1 x^{(i)} - y^{(i)})
\right]
\\
&=
\frac{1}{m} \sum_{i = 0}^{m}
\left[
\beta_0 x^{(i)}+ \beta_1 {x^{(i)}}^2 - x^{(i)}y^{(i)}
\right]
\\
&=
\beta_0 \frac{1}{m} \sum_{i = 0}^{m} x^{(i)} +
\beta_1 \frac{1}{m} \sum_{i = 0}^{m}  {x^{(i)}}^2 -
\frac{1}{m} \sum_{i = 0}^{m} x^{(i)}y^{(i)}
\\
&=
\beta_0 \bar x +
\beta_1 \bar{x^2} -
\bar {xy}
\end{align*}

## Linear Regression by Gradient Decent
```{r Problem Statement 2, include = F}
# Linear regression by gradient descent in R
# The purpose of this exercise is to test what is above and implement that in R.
# Steps:
# a) Generate some random data for X and Y (use rnorm function in R)
# b) Check linearity by running lm(y~x) and get the coefficient
# c) Build Cost and gradient descent function
# d) The output of betas (ğ›½0  + ğ›½1) should be equal to the coefficients above
```

```{r}
# Generates linear data with normal residuals
set.seed(123)
x <- rnorm(n = 30)

epsilon <- rnorm(n = 30)

y <- 5*x + 1 + epsilon

plot(x,y)
```

```{r}
ols_cost <- function(x, y, beta_0, beta_1){
    cost <- (y - (beta_1 * x + beta_0))^2
    return(sum(cost))
}
```

```{r}
slr_gradient_desc <- function(x, y, alpha){

    # Summary statistic calculations.
    # Helps to calculate the gradient faster.
    x_bar <- mean(x)
    y_bar <- mean(y)
    xy_bar <- mean(x*y)
    x_sqbar <- mean(x^2)

    # initial guess for beta_0 and beta_1
    beta_0 <- y_bar
    beta_1 <- 0

    cost_0 <- ols_cost(x, y, beta_0, beta_1)

    # Dataframe to store results.
    df_results <- data.frame(beta_0 = c(beta_0),
                            beta_1 = c(beta_1),
                            error = c(cost_0))

    # A counter to determine is the error is unchanging.
    count_same <- 0

    while(count_same < 10){

        # Cost prior to beta adjustment.
        cost_start <- ols_cost(x, y, beta_0, beta_1)

        #print(cost_start)

        print(c(beta_0, beta_1))

        # Calculate gradient values.
        g_0 <- beta_0 + (beta_1 * x_bar) - y_bar
        g_1 <- (beta_0 * x_bar) + (beta_1 * x_sqbar) - xy_bar

        # Update betas.
        beta_0 <- beta_0 + (alpha * g_0)
        beta_1 <- beta_1 + (alpha * g_1)

        # Calculate new cost.
        cost_after <- ols_cost(x, y, beta_0, beta_1)

        #print(cost_after)

        print(c(beta_0, beta_1))

        # Check cost relation.
        #if(cost_start < cost_after) {
            #return("Bad alpha, over shot minimum! Lower alpha and try again")
        if(cost_start == cost_after) {
            df_results <- cbind(df_results,
                                c(beta_0, beta_1, cost_after))
            count_same <- count_same + 1
        } else {
            df_results <- cbind(df_results,
                                c(beta_0, beta_1, cost_after))
        }
    }
    return(df_results)
}
```

## Test
```{r}
slr_gradient_desc(x, y, alpha = 0.01)
```

```{r}
beta_0 <- y_bar
beta_1 <- 0

for(i in 10000){
    # Summary statistic calculations.
    # Helps to calculate the gradient faster.
    x_bar <- mean(x)
    y_bar <- mean(y)
    xy_bar <- mean(x*y)
    x_sqbar <- mean(x^2)

    # initial guess for beta_0 and beta_1

    g_0 <- beta_0 + (beta_1 * x_bar) - y_bar
    g_1 <- (beta_0 * x_bar) + (beta_1 * x_sqbar) + xy_bar

    # Update betas.
    beta_0 <- beta_0 + (0.1 * g_0)
    beta_1 <- beta_1 + (0.1 * g_1)

    print(c(beta_0, beta_1, ols_cost(x, y, beta_0, beta_1)))
}
```


## Linear Model on Airbnb Data
